Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-06 paper.tex
add figures

% note on the disjoint set of intervals for each lineage under the SMC
As noted by \cite{mcvean_approximating_2005}, the backwards-in-time description of the 
SMC is slightly different from the left-to-right approach in that recombination 
in non-ancestral material is possible. Although this does not affect the distribution 
of marginal genealogies, it does mean that not all lineages in $L$ can be represented 
by a single interval. More precisely, all lineages will be associated with 
a single interval up to the point where one section of the 
genome has reached its most recent common ancestor but the neighbouring parts have not.

@@

Recording the full process as described above will result in unary nodes.
These are nodes that will have only one child in one or more local trees 
(see fig. \ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages hit by a recombination event. 
In case of a coalescence event, a node is registered along the full 
extent of the original lineage $X_c$. 
Coalescence between two lineages without perfectly 
overlapping intervals will therefore result in locally unary nodes.\\

@@

Here we will compute the likelihood for an ARG
under the SMC for which we have no information on the time to any 
of the recombination events. We will assume however, that the identity 
of the lineage involved in each recombination event can be correctly 
inferred by relying on the unary nodes (see section X.X on locally unary nodes). 
This enables us to extend likelihood computations to the 
a more general class of ARGs.

\begin{figure}[h!] \label{fig:smc-unary}
\centering
\includegraphics[width=\textwidth]{smc-custom-haplotypes-mini.png}
\caption{SMC with unary nodes}
\end{figure}

@@

The SMC with unary nodes
Unary nodes connect marginal trees to the ARG.
These are inferred by \tsinfer, \kwarg, \argweaver. %add citation

% what happens in the partial or complete absence of unary nodes
% even in the absence of unary nodes we can compute a likelihood, however, this has an impact.

% The SMC and unary nodes
% ADD: figure here that tries to deal with all possible causes of confusion
% unary nodes mainly

% highlight the amount of information that needs to be tracked to compute a likelihood
% for the exact Hudson CwR

@@

\begin{figure}[h!] \label{fig:algo}
\centering
\includegraphics[width=0.9\textwidth]{smc_algo.png}
\caption{Description of algorithm}
\end{figure}

\begin{figure}[h!] \label{fig:algo-flat}
\centering
\includegraphics[width=\textwidth]{smc_algo_flat.png}
\caption{Description of algorithm}
\end{figure}

@@

Although the argument outlined in the previous paragraph is valid for any ARG, here we 
will describe how to (efficiently) extract the information required to associate a
tree sequence encoding of an ARG with a likelihood under the SMC.\\

Formally, we represent a tree sequence using a set of tables \citep{kelleher_efficient_2018}. 
This simple tabular representation of any ARG $\mathcal{G} = (N, E)$, can be 
augmented with additional tables with information on sequence variation
using a site and mutation table. The node table stores information on all 
haplotypes $N$ in the tree sequence. Information about how nodes relate to 
each other along the genome is defined in a tabular encoding of $E$.
Each row in this edge table constitutes a tuple \texttt{(left, right, parent, child)}.
Note that we defined the extent of any edge $\epsilon \in \mathcal{G}$ as the union of 
disjoint intervals. In the tree sequence encoding however, $\epsilon$ will be represented 
by a single entry for each contiguous interval. As remarked before, under the SMC, this 
distinction only matters beyond the root of the genealogy. The distinction between both 
types of edges will therefore not be made in the remainder of the paper.\\

The likelihood computation as described above can be done with a single pass 
of the edge table.
The central problem however is to efficiently maintain the state $I_e(t)$, which counts 
the number of lineages (the lineage associated with) the focal edge $\epsilon$ can coalesce 
with at time $t$.

The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
that we can impose a strict total order on all edges based on their starting point.

Because of the SMC:
an edge starting at $x$ can only coalesce 
with a subset of the other edges that make up the local tree at position $x$.

Because local tree changes only occur whenever edges are removed and/or inserted,
we will rely on the ordering of these insertions and removals as implemented in
\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\

@@

By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence.
Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
benefit from this. %elaborate
We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).
Similarly, any subset of the tree sequence that is itself a valid tree sequence can again 
be used to compute a likelihood. Such operations include: slicing genome both in time, and space.

\subsubsection{Slicing the ARG in time}
% TO DO: give example why this might be useful.

@@

\section{Discussion}
% summary: what did we do here

@@

First step towards a genealogy sampler (for any ARG, and at scale): quantifying uncertainty 
(with inferred ARG as starting point)

resolve polytomies in a model-based way

>>>>

% note on the disjoint set of intervals for each lineage under the SMC
As noted by \cite{mcvean_approximating_2005}, the backwards-in-time description of the 
SMC is slightly different from the left-to-right approach in that recombination 
in non-ancestral material is possible (see fig.\ref{fig:smc-unary}). 
Although this does not affect the distribution 
of marginal genealogies, it does mean that not all lineages in $L$ can be represented 
by a single interval. More precisely, all lineages will be associated with 
a single interval up to the point where one section of the 
genome has reached its most recent common ancestor but the neighbouring parts have not.

@@

Recording the full process as described above will result in unary nodes.
These are nodes that will have only one child in one or more local trees 
(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages hit by a recombination event. 
In case of a coalescence event, a node is registered along the full 
extent of the original lineage $X_c$. 
Coalescence between two lineages without perfectly 
overlapping intervals will therefore result in locally unary nodes.\\

@@

Here we will compute the likelihood for an ARG
under the SMC for which we have no information on the time to any 
of the recombination events. 
We will assume however, that the identity 
of the lineage involved in each recombination event can be correctly 
inferred by relying on the unary nodes (see section X.X on locally unary nodes).
As highlighted in fig.\ref{fig:smc-unary}, unary nodes can be inferred from the 
available haplotype information. 
Currently, the following methods return this information by default 
as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome_wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.
The presence of these unary nodes enables us to extend likelihood computations 
to the a more general class of ARGs.

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc-custom-haplotypes-mini.png}
\caption{SMC with unary nodes: Tree sequence encoding of an ARG (top left) 
generated under the SMC. The haplotypes that make up the lineages coalescing 
at the root of each marginal tree are highlighted at the top. This ancestral 
material is coloured based on the samples that inherit from it.
These reveal the details on the presence of the unary nodes. 
Note how the presence of the unary nodes 
modifies the left-to-right logic of the prune regrafts 
under the SMC (blue dotted arrows).
The unary nodes allow us to keep track of the full extent of all lineages 
through time (under the SMC) and thus infer the lineage along which a 
recombination event happened.}
\label{fig:smc-unary}
\end{figure}

@@

Recording/inferring unary nodes allow us to completely reconstruct 
$X(t)$ through time under the SMC up to the exact 
time of the recombination events. \\

The SMC and unary nodes. \\

inferring recombination = change of parent. \\

@@

\begin{figure}[h!] \label{fig:algo}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_algo.png}
\caption{Description of algorithm}
\end{figure}

\begin{figure}[h!] \label{fig:algo-flat}
\centering
\includegraphics[width=\textwidth]{figures/smc_algo_flat.png}
\caption{Description of algorithm}
\end{figure}

@@

Although the argument outlined in the previous paragraph is valid for any ARG, here we 
will describe how to (efficiently) extract the information required to associate a
tree sequence encoding of an ARG with a likelihood under the SMC (see fig. \ref{fig:smc-unary}).\\

Formally, we represent a tree sequence using a set of tables \citep{kelleher_efficient_2018}. 
This simple tabular representation of any ARG $\mathcal{G} = (N, E)$, can be 
augmented with additional tables with information on sequence variation
using a site and mutation table. The node table stores information on all 
haplotypes $N$ in the tree sequence. Information about how nodes relate to 
each other along the genome is defined in a tabular encoding of $E$.
Each row in this edge table constitutes a tuple \texttt{(left, right, parent, child)}.
Note that we defined the extent of any edge $\epsilon \in \mathcal{G}$ as the union of 
disjoint intervals. In the tree sequence encoding however, $\epsilon$ will be represented 
by a single entry for each contiguous interval. As remarked before, under the SMC, this 
distinction only matters beyond the root of the genealogy. The distinction between both 
types of edges will therefore not be made in the remainder of the paper.
% Although it does requires us to keep track of some additional information to avoid
% wrongfully assuming a recombination event happened.
% Similarly, the distinction between the lineage associated with an edge, and the edge 
% itself can render things very verbose.
\\

The likelihood computation as described above can be done with a single pass 
of the edge table.
The central problem however is to efficiently maintain the state $I_e(t)$, which counts 
the number of lineages (the lineage associated with) the focal edge $\epsilon$ can coalesce 
with at time $t$. 
The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
that we can impose a strict total order on all (overlapping) lineages based on their 
starting point.

%%%% improve this %%%%%%5
The lineages associated with a particular edge can only coalesce with other lineages 
an edge starting at $x$ can only coalesce 
with a subset of the other edges that make up the local tree at position $x$.
%%%% 
Because local tree changes only occur whenever edges are removed and/or inserted,
we will rely on the ordering of these insertions and removals as implemented in
\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\

@@

By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence.

The \texttt{tskit} library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
and simplification.
Similarly, any subset of the tree sequence that is itself a valid tree sequence can again 
be used to compute a likelihood. Such operations include: slicing genome both in time, and space.

Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
benefit from this.

We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).

% HOW DO WE HIGHLIGHT THESE FEATURES: SUPPLEMENTS?
\subsubsection{Slicing the ARG in time}
% TO DO: give example why this might be useful.

@@

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. 
Important step moving forwards towards bridging the gap between the tremendous 
recent advances in ARG inference and current state of the art statistical inference of 
past demography and selection (but see). % some/many of these approaches 
% ignore recombination

@@

First step towards a genealogy sampler (for any ARG, and at scale): 
quantifying uncertainty 
(with inferred ARG as starting point). Compared to existing MCMC approaches, 
we further restrict the size of ARG space by 
integrating out the exact time to the recombination event.

% discuss accurracy:
% there will be scenarios where the assumptions made here are not met: e.g. polytomies
% yet we can still always compute a likelihood.
% we can alwyas associate a value, what remains to be verified what the information is in that
% case.
